---
output:
  github_document:
    fig_width: 3
---

[![Coverage status](https://codecov.io/gh/martakarass/adept/branch/master/graph/badge.svg)](https://codecov.io/github/martakarass/adept?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/adept)](https://CRAN.R-project.org/package=adept)
[![](https://cranlogs.r-pkg.org/badges/grand-total/adept)](https://cran.r-project.org/package=adept)
[![](https://cranlogs.r-pkg.org/badges/last-month/adept)](https://cran.r-project.org/package=adept)
[![R build status](https://github.com/martakarass/adept/workflows/R-CMD-check/badge.svg)](https://github.com/martakarass/adept/actions)
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

### Overview

The `adept` package implements ADaptive Empirical Pattern Transformation (ADEPT) method^[Karas, M., Straczkiewicz, M., Fadel, W., Harezlak, J., Crainiceanu, C., Urbanek, J.K. *Adaptive empirical pattern transformation (ADEPT) with application to walking stride segmentation*, Submitted to *Biostatistics*, 2018.] for pattern segmentation from a time-series. ADEPT is optimized to perform fast, accurate walking strides segmentation from high-density data collected with a wearable accelerometer during walking. The method was validated using data collected with sensors worn at left wrist, left hip and both ankles. 

### Installation

Install `adept` package from GitHub. 

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("martakarass/adept")
```

### Example 1

We simulate a time-series `x`. We assume that `x` is collected at a frequency of 100 Hz, there is one shape of a pattern within `x`, each pattern lasts 1 second, and there is no noise in collected data. 

```{r, fig.width=10, fig.height=4}
true.pattern <- cos(seq(0, 2 * pi, length.out = 100))
x <- c(true.pattern[1], replicate(10, true.pattern[-1]))

par(mfrow = c(1,2), cex = 1)
plot(true.pattern, type = "l", xlab = "", ylab = "", main = "Pattern")
plot(x, type = "l", xlab = "", ylab = "", main = "Time-series x")
```

We segment pattern from data. We assume that a perfect template is available. We use a grid of potential pattern durations of {0.9, 0.95, 1.03, 1.1} seconds; the grid is imperfect in a sense it does not contain the duration of the true pattern used in `x` simulation. 

```{r, fig.width=2.5, fig.height=2.3}
library(adept)

segmentPattern(
  x = x,
  x.fs = 100,
  template = true.pattern,
  pattern.dur.seq = c(0.9, 0.95, 1.03, 1.1),
  similarity.measure = "cor",
  compute.template.idx = TRUE)
```

The segmentation result is a data frame, where each row describes one identified pattern occurrence:

* `tau_i` - index of `x` where pattern starts,
* `T_i` - pattern duration, expressed in `x` vector length,
* `sim_i` - similarity between a template and `x`,
* `template_i` - index of a template best matched to a time-series `x` (here: one template was used, hence all `template_i`'s equal 1).

We then assume a grid of potential pattern durations which contains the duration of the true pattern used in data simulation. A perfect match (`sim_i = 1`) between a time-series `x` and a template is obtained. 

```{r, fig.width=2.5, fig.height=2.3}
segmentPattern(
  x = x,
  x.fs = 100,
  template = true.pattern,
  pattern.dur.seq = c(0.9, 0.95, 1, 1.03, 1.1),
  similarity.measure = "cor",
  compute.template.idx = TRUE)
```

### Example 2

We simulate a time-series `x`. We assume that `x` is collected at a frequency of 100 Hz, there are two shapes of a pattern within `x`, patterns have various duration, and there is no noise in collected data. 

Then, we generate `x2` as a noisy version of `x`. 

```{r, fig.width=10, fig.height=4}
true.pattern.1 <- cos(seq(0, 2 * pi, length.out = 200))
true.pattern.2 <- true.pattern.1
true.pattern.2[70:130] <- 2 * true.pattern.2[min(70:130)] + abs(true.pattern.2[70:130])
x <- numeric()
for (vl in seq(70, 130, by = 10)){
  true.pattern.1.s <- approx(
    seq(0, 1, length.out = 200), 
    true.pattern.1, xout = seq(0, 1, length.out = vl))$y
  true.pattern.2.s <- approx(
    seq(0, 1, length.out = 200), 
    true.pattern.2, xout = seq(0, 1, length.out = vl))$y
  x <- c(x, true.pattern.1.s[-1], true.pattern.2.s[-1])
  if (vl == 70) x <- c(true.pattern.1.s[1], x)
}
set.seed(1)
x2 <- x + rnorm(length(x), sd = 0.5)

par(mfrow = c(1,2), cex = 1)
plot(true.pattern.1, type = "l", xlab = "", ylab = "", main = "Pattern 1")
plot(true.pattern.2, type = "l", xlab = "", ylab = "", main = "Pattern 2")
par(mfrow = c(1,1), cex = 1)
plot(x, type = "l", xlab = "", ylab = "", main = "Time-series x")
plot(x2, type = "l", xlab = "", ylab = "", main = "Time-series x2")
```

We segment `x`. We assume a perfect grid of potential pattern duration, {0.7, 0.8, 0.9, 1.0, 1.1, 1.2, 1.3} seconds. 
```{r}
segmentPattern(
  x = x,
  x.fs = 100,
  template = list(true.pattern.1, true.pattern.2),
  pattern.dur.seq = seq(0.7, 1.3, by = 0.1),
  similarity.measure = "cor",
  compute.template.idx = TRUE)
```

We segment `x2`.  

```{r}
segmentPattern(
  x = x2,
  x.fs = 100,
  template = list(true.pattern.1, true.pattern.2),
  pattern.dur.seq = seq(0.7, 1.3, by = 0.1),
  similarity.measure = "cor",
  compute.template.idx = TRUE)
```

We now use `x.adept.ma.W` argument to smooth `x2` before similarity matrix computation in the segmentation procedure (see `?segmentPattern` for details). We also assume a more dense grid of potential pattern duration. We observe that `sim_i` values obtained are higher than in the previous segmentation case. 

```{r, fig.width=10, fig.height=4}
par(mfrow = c(1,1), cex = 1)
plot(windowSmooth(x = x2, x.fs = 100, W = 0.1), 
     type = "l", xlab = "", ylab = "", main = "Time-series x2 smoothed")
segmentPattern(
  x = x2,
  x.fs = 100,
  template = list(true.pattern.1, true.pattern.2),
  pattern.dur.seq = 70:130 * 0.01,
  similarity.measure = "cor",
  x.adept.ma.W = 0.1,
  compute.template.idx = TRUE)
```


### Vignettes 

Vignettes are available to better demonstrate package methods usgae. 

1. Vignette [Introduction to adept package](https://martakarass.github.io/adept/articles/adept-intro.html) introduces ADEPT algorithm and demonstrates the usage of `segmentPattern` function which implements ADEPT approach. Here, we focus on examples with simulated data.

2. Vignette [Walking strides segmentation with adept](https://martakarass.github.io/adept/articles/adept-strides-segmentation.html) provides an example of segmentation of walking strides (two consecutive steps) in sub-second accelerometry data with `adept` package. The exemplary dataset is a part of the `adeptdata` package. We demonstrate that ADEPT can be used to perform automatic and precise walking stride segmentation from data collected during a combination of running, walking and resting exercises. We introduce how to segment data: 

    1. with the use of stride templates that were pre-computed based on data from an external study (attached to `adeptdata` package), 
    2. by deriving new stride templates in a semi-manual manner. 
    
### References
