---
title: "Marta Karas"
author: "Introduction to `adept` package"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r package_options, include=FALSE}
knitr::opts_knit$set(global.par = TRUE)
```

`adept` package implements ADaptive Empirical Pattern Transformation (ADEPT) method for pattern segmentation
from a time-series `x`. ADEPT was designed with the aim of performing fast, accurate walking strides segmentation from high-density data collected from wearable accelerometer worn during continuous walking activity. 

# ADEPT method   

ADEPT identifies pattern occurrenes from a time-series `x` via maximizing similarity (correlation, covariance etc.) between time-series `x` windows and a pattern template(s). 

To address a possible scenario in which a pattern occurrence is changing its duration within time-series `x`, for each window of `x`, a pattern template(s) is scaled to various scale parameters to allow for obtaining higher similarity between `x` and a template. Also, multiple pattern templates are allowed simultaneously to address that a pattern may potentially be changing its shape over the time-series `x`. 

In practice, a pre-defined pattern template may be derived as an empirical pattern, that is, data-derived vector representing a pattern of interest.

# Pattern segmentation with `adept` package  

## Install / load `R` packages as needed

- Install (if not installed yet)

```{r, eval = FALSE}
# install.packages("devtools")       ## for installing packages from GitHub repository
```

- Install `adept` from GitHub (need `devtools` package being installed to do it)

```{r, eval = FALSE}
devtools::install_github("martakarass/adept")
```

- Load packages

```{r}
library(adept)
```

## Examples 1: simulated signals, same pattern length, no noise 

### Example 1(a) 

Simulated data: 

- No noise in simulated signal `x`
- all simulated pattern occurences of the same length (vector length: `101`)

```{r, fig.width=8, fig.height=2.7}
## Simulate signal and template
x0 <- cos(seq(0, 2 * pi * 10, length.out = 1001))
x  <- x0
template <- x0[1:101]

## Plot signal
par(mar = c(4.1, 4.1, 2.1, 2.1)) 
plot(x, type = "l", main = "", xlab = "time-series x index")
```

- Plot pattern template

```{r, fig.width=3.0, fig.height=2.7}
plot(template, type = "l", main = "", xlab = "Duration phase", ylab = "Template", col = "red")
```



















